---
title: "DA401 Suicide Risk"
author: "Jiayu Zhao"
date: "11/15/2022"
output:
  html_document:
    code_folding: hide
    theme: journal
    toc: yes
    toc_depth: 3
  pdf_document:
    latex_engine: xelatex
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(ggplot2)
library(caret)
library(MASS)
library(jtools)
library(nnet)
library(interactions)
library(keras)
library(tensorflow)
library(mltest)

setwd("~/Documents/GitHub/DA401_SuicideRisk_research")
load("NSDUH_2020.RData")
load("NSDUH_2019.RData")
load("NSDUH_2018.RData")
```

## Data cleaning
```{r}
true20<- NSDUH_2020%>%
  filter(auinpyr == 1|auoptyr == 1)
true19<-PUF2019_100920 %>%
  filter(auinpyr == 1|auoptyr == 1)
true18<-PUF2018_100819 %>%
  filter(auinpyr == 1|auoptyr == 1)
true18$QUESTID2<-as.integer(true18$QUESTID2)
true19$QUESTID2<-as.integer(true19$QUESTID2)
SuicideRisk<-bind_rows(true20, true19, true18)

SR<-SuicideRisk %>% 
  dplyr::select(DSTHOP30, mhsuithk, mhsuipln, mhsuitry, K6SCMON, SMIPP_U, AUNMPSY2, AUNMPGE2, AUNMMED2,AUNMAHS2,
                AUNMRES2,AUNMSFA2, AUNMMEN2, AUNMTHE2, AUNMDOC2, AUNMCLN2, AUNMDTM2, AUNMOTO2, amdelt, ASDSOVL2,
                CATAG6, irmarit,irsex, IREDUHIGHST2, NEWRACE2, wrkdpstyr, income, ALCWD2SX, alcndmor)

val1 = c(83,91,93,94,97,98,99) #value for the invalid data (unknow, no repsonse...)
SR <- as.data.frame(sapply(SR, function(x) replace(x, x %in% val1, 0)))
SR$DSTHOP30=as.numeric(factor(SR$DSTHOP30, levels = c(5,4,3,2,1), labels = c(1,2,3,4,5)))

val2 = c(985,994,997,998,999)
SR[7:18] <- sapply(SR[7:18], function(x) replace(x, x %in% val2, 0))

#Combine treatment from multiple facilities into a single var
SR$inpatient<-rowSums(SR[7:12])
SR$outpatient <- rowSums(SR[13:18])
```

## Risk Index

#### Suicide Risk Level (categorical)

```{r}
SR<-SR %>% mutate(risk_level = case_when(as.numeric(DSTHOP30>3) + as.numeric(mhsuithk==1)+ 
                                           as.numeric(mhsuipln == 1) + as.numeric(mhsuitry == 1) +
                                           as.numeric(K6SCMON > 16) + as.numeric(SMIPP_U > 0.65) >= 3 ~"high",
                                         as.numeric(DSTHOP30<4 & DSTHOP30>2) + as.numeric(mhsuithk == 1) +
                                           as.numeric(mhsuipln == 1 | mhsuitry == 1) + 
                                           as.numeric(K6SCMON <= 16 & K6SCMON >= 8) + 
                                           as.numeric(SMIPP_U <= 0.65 & SMIPP_U > 0.3) >= 2 ~"moderate",
                                         as.numeric(DSTHOP30<2) + as.numeric(K6SCMON < 8) +
                                           as.numeric(mhsuithk== 1 | mhsuipln == 1 | mhsuitry == 1) +
                                           as.numeric(SMIPP_U <= 0.3) >= 2 ~ "low"))
SR$risk_level[is.na(SR$risk_level)] <- "no"
```

#### Suicide Risk Score (continuous)
```{r}
SR_reg<-SR
SR_reg<-SR_reg[c(-7:-18)]
SR_reg[c(7:17)] <- lapply(SR_reg[c(7:17)], factor)

SR_reg$K6SCMON = as.numeric(scale(SR_reg$K6SCMON))
SR_reg <- SR_reg %>% rowwise() %>%
  mutate(risk_score = sum(c_across(DSTHOP30:SMIPP_U)))
```


## Regression

### ANOVA
```{r}
library(tidyverse)
library(ggpubr)
library(rstatix)

set.seed(123)
SR_reg %>% sample_n_by(., size = 1)

# Check normaility assumption
normality <- lm(risk_score ~ amdelt+ASDSOVL2+CATAG6+irmarit+irsex+IREDUHIGHST2+NEWRACE2+wrkdpstyr+
                  income+alcndmor+ALCWD2SX+inpatient+outpatient, data = SR_reg)
# Create a QQ plot of residuals
ggqqplot(residuals(normality))
# Compute Shapiro-Wilk test of normality
shapiro_test(residuals(normality))

## Homogeneity of variance
SR_reg %>% levene_test(risk_score ~ amdelt+ASDSOVL2+CATAG6+irmarit+irsex+IREDUHIGHST2+NEWRACE2+wrkdpstyr+
                         income+alcndmor+ALCWD2SX+inpatient+outpatient, data = SR_reg)
```


```{r}
# Test
aov<-aov(data = SR_reg, risk_score ~ (amdelt+CATAG6+irmarit+irsex+IREDUHIGHST2+NEWRACE2+wrkdpstyr+
                                         income+alcndmor+ALCWD2SX+inpatient+outpatient)^2)
summary(aov)

#Post-hoc test
post_hoc1<-TukeyHSD(aov1, which = c('amdelt','CATAG6','irmarit','irsex','IREDUHIGHST2','NEWRACE2','income','ALCWD2SX','alcndmor')) 
post_hoc1
```


### Plots for Interactive Effects

##### Linear models for two-way interactions
```{r}
## MDE
mde_age<-glm(data = SR_reg, risk_score ~ amdelt*CATAG6)
mde_mar<-glm(data = SR_reg, risk_score ~ amdelt*irmarit)
mde_sex<-glm(data = SR_reg, risk_score ~ amdelt*irsex)
mde_edu<-glm(data = SR_reg, risk_score ~ amdelt*IREDUHIGHST2)
mde_emp<-glm(data = SR_reg, risk_score ~ amdelt*wrkdpstyr)
mde_income<-glm(data = SR_reg, risk_score ~ amdelt*income) #!
mde_tolerance<-glm(data = SR_reg, risk_score ~ amdelt*alcndmor) #!
mde_in<-lm(data = SR_reg, risk_score ~ amdelt*inpatient)
```

```{r}
## Age
age_sex<-glm(data = SR_reg, risk_score ~ CATAG6*irsex)
age_edu1<-glm(data = SR_reg, risk_score ~ IREDUHIGHST2*CATAG6)
age_emp<-glm(data = SR_reg, risk_score ~ CATAG6*wrkdpstyr)

## Marital
mar_emp<-glm(data = SR_reg, risk_score ~ irmarit*wrkdpstyr)
mar_tolerance<-glm(data = SR_reg, risk_score ~ irmarit*alcndmor)
mar_income<-glm(data = SR_reg, risk_score ~ irmarit*income) 
mar_in<-lm(data = SR_reg, risk_score ~ irmarit*inpatient)
mar_symp<-glm(data = SR_reg, risk_score ~ irmarit*ALCWD2SX)
```

```{r}
## Sex
sex_edu<-glm(data = SR_reg, risk_score ~ irsex*IREDUHIGHST2)

## Education
edu_in<-lm(data = SR_reg, risk_score ~ IREDUHIGHST2*inpatient)
edu_out<-lm(data = SR_reg, risk_score ~ IREDUHIGHST2*outpatient) 
edu_tolerance<-glm(data = SR_reg, risk_score ~ IREDUHIGHST2*alcndmor)

## Race
race_in<-lm(data = SR_reg, risk_score ~ NEWRACE2*inpatient)

## Employment
emp_income<-glm(data = SR_reg, risk_score ~ wrkdpstyr*income) #!
emp_in<-lm(data = SR_reg, risk_score ~ wrkdpstyr*inpatient) #!

## Income
income_tolerance<-glm(data = SR_reg, risk_score ~ income*alcndmor) #!
income_in<-lm(data = SR_reg, risk_score ~ income*inpatient)

## Symptoms
symp_out<-lm(data = SR_reg, risk_score ~ ALCWD2SX*outpatient) #!
```

##### Plot the interactive effective
```{r}
## MDE
mde_ageP<-cat_plot(mde_age, pred = amdelt, modx = CATAG6, geom = "line") #No, because the lines are parallel 
mde_marP<-cat_plot(mde_mar, pred = amdelt, modx = irmarit, geom = "line") #No
mde_sexP<-cat_plot(mde_sex, pred = amdelt, modx = irsex, geom = "line") #No
mde_eduP<-cat_plot(mde_edu, pred = IREDUHIGHST2, modx = amdelt, geom = "line") #No
mde_empP<-cat_plot(mde_emp, pred = amdelt, modx = wrkdpstyr, geom = "line") #No
mde_incomeP<-cat_plot(mde_income, pred = amdelt, modx = income, geom = "line") #No
mde_toleranceP<-cat_plot(mde_tolerance, pred = amdelt, modx = alcndmor, geom = "line") #No
mde_inP<-interact_plot(mde_in, pred = inpatient, modx = amdelt, legend.main = "MDE", 
                        x.label = "# of Times Receiving Inpatient Treatment",
                        modx.labels = c("Yes","No"),
                        main.title = "Interactions between MDE and Inpatient Treatment")
```

```{r}
## Age
age_sexP<-cat_plot(age_sex, pred = CATAG6, modx = irsex, geom = "line")  #No
age_empP<-cat_plot(age_emp, pred = CATAG6, modx = wrkdpstyr, geom = "line") #No
age_eduP<-cat_plot(age_edu, pred = IREDUHIGHST2, modx = CATAG6,geom = "line",
                   legend.main = "Age",
                   x.label = "Education Level",
                   modx.labels = c("18-25","26-34","35-49","50-64","65 or older"),
                   pred.labels = c("5th grade or less","6th grade","7th grade","8th grade","9th grade",
                                   "10th grade","11th/12th grade, no diploma","High school diploma",
                                   "College credit, no degree","Associate's degree",
                                   "College graduate or higher"),
                   main.title = "Interactions between Education Level and Age")+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

```{r}
## Marital
mar_empP<-cat_plot(mar_emp, pred = irmarit, modx = wrkdpstyr, geom = "line")  #No
mar_drinkP<-cat_plot(mar_drink, pred = irmarit, modx = alcndmor, geom = "line")  #No
mar_incomeP<-cat_plot(mar_income, pred = income , modx = irmarit, geom = "line", legend.main = "Marital Status",
                      x.label = "Income Level",
                      pred.labels = c("Less than $20,000","$20,000-$49,999","$50,000-$74,999","$75,000 or More"),
                      modx.labels = c("Married","Widowed","Divorced or seperated","Never been married"),
                      main.title = "Interactions between Income Level and Marital Status") 
mar_inP<-interact_plot(mar_in, pred = inpatient , modx = irmarit) #No
mar_sympP<-cat_plot(mar_symp, pred = irmarit , modx = ALCWD2SX, geom = "line",legend.main = "Two or more \nwithdrawal symptoms",
                    pred.labels = c("Married","Widowed","Divorced or seperated","Never been married"),
                    main.title = "Interactions between Withdrawal Symptoms and Marital Status",
                    x.label = "Marital Status",
                    modx.labels = c(0,"Yes","No"))+
  theme(axis.text.x = element_text(angle = 30, vjust = 1, hjust=1))
```

```{r}
## Education
edu_inP<-interact_plot(edu_in, pred = inpatient, modx = IREDUHIGHST2, colors="Rainbow",
                       legend.main = "Highest Education Level",
                       x.label = "# of Times Receiving Inpatient Treatment",
                       modx.labels = c("5th grade or less","6th grade","7th grade","8th grade","9th grade",
                                       "10th grade","11th/12th grade completed, \nno diploma","High school diploma",
                                       "College credit, no degree","Associate's degree",
                                       "College graduate or higher"),
                       main.title = "Interactions: \nEducation & Inpatient Treatment") +
  theme(axis.title.x=element_text(size=7), plot.title = element_text(size = 10), legend.text = element_text(size = 6), 
        legend.title = element_text(size = 8), aspect.ratio=2/1) #yes
edu_outP<-interact_plot(edu_out, pred = outpatient, modx = IREDUHIGHST2, colors="Rainbow",
                        legend.main = "Highest Education Level",
                        modx.labels = c("5th grade or less","6th grade","7th grade","8th grade","9th grade",
                                       "10th grade","11th/12th grade completed, \nno diploma","High school diploma",
                                       "College credit, no degree","Associate's degree",
                                       "College graduate or higher"),
                        x.label = "# of Times Receiving Outpatient Treatment",
                        main.title = "Interactions: \nEducation & Outpatient Treatment") +
  theme(axis.title.x=element_text(size=7), plot.title = element_text(size = 10), legend.text = element_text(size = 6), 
        legend.title = element_text(size = 8), aspect.ratio=2/1)#yes
```

```{r}
## Sex
sex_eduP<-cat_plot(sex_edu, modx = irsex, pred = IREDUHIGHST2, geom = "line") #No

## Race
race_inP<-interact_plot(race_in, pred = inpatient, modx = NEWRACE2, geom = "line",legend.main = "Race",
                        x.label = "# of Times Receiving Inpatient Treatment",
                        main.title = "Interactions between Income Level and Marital Status",
                        modx.labels = c("NonHisp White","NonHisp Black/Afr Am","NonHisp Native Am/AK Native",
                                        "NonHisp Native HI/Other Pac Isl","NonHisp Asian","NonHisp more than one race",
                                        "Hispanic"))

## Employment
emp_incomeP<-cat_plot(emp_income, modx = wrkdpstyr, pred = income, geom = "line") #No
emp_inP<-interact_plot(emp_in, pred = inpatient, modx = wrkdpstyr, geom = "line") #no

## Income
income_toleranceP<-cat_plot(income_tolerance, pred = income, modx = alcndmor, geom = "line") #no
income_inP<-interact_plot(income_in, pred = inpatient, modx = income, geom = "line") #no

## Symptoms
symp_outP<-interact_plot(symp_out, pred = outpatient, modx = ALCWD2SX, geom = "line", 
                         legend.main = "Two or more \nwithdrawal symptoms",
                         modx.labels = c(0,"Yes","No"),
                         x.label = "# of Times Receiving Outpatient Treatment",
                         main.title = "Interactions between Withdrawal Symptoms and Outpatient Treatment")
```


```{r}
#Figure 1: interaction between outpatient treatment and MDE
mde_outP
```

```{r}
#Figure 2: Interaction between education and treatment
edu<-ggarrange(edu_inP+theme(legend.margin = margin(t = 0, unit='cm')), 
               edu_outP+theme(legend.margin = margin(t = 0, unit='cm')), 
               nrow=1, common.legend = TRUE, legend="right")
edu
```

```{r}
#Fig 3: interaction between outpatient and withdrawal symptoms
symp_outP
```

```{r}
#Figure 4: interaction between age and education
age_eduP
```

```{r}
#Figure 5: Marital status interactions
mar_incomeP
mar_sympP
```



## Classification
```{r}
SR<-SR[c(-1:-18)]
SR<-na.omit(SR[c(1,3:14)]) #omit the variable with too much NA and all rows w/ NA (drop ASDSOVL2)
SR[c(1:10)] <- lapply(SR[c(1:9)], factor)
```

#### Training Testing set
```{r}
# Training Testing set
set.seed(1)
index = createDataPartition(y=SR$risk_level, p=0.7, list=FALSE) 
sr_train =SR[index,]
sr_test = SR[-index,]

ctrl = trainControl(method="repeatedcv", number = 10, repeats = 5)
```

#### knn
```{r}
train_knn<-sr_train
#train_knn$risk_level = factor(train_knn$risk_level, levels = c(1,2,3,4), labels = c("no","low","moderate","high"))
trctrl_knn= trainControl(method = "repeatedcv", number = 10, repeats = 5, classProbs = TRUE)
knn = caret::train(risk_level ~ ., data = train_knn, method = 'knn', trControl=trctrl_knn,
            preProcess = c("center", "scale"),tuneLength = 25)
#plot(knn)
```

#### Decision Tree
```{r}
ctrl_dt = trainControl(method="repeatedcv",number=10, repeats=5)

dtree = caret::train(risk_level~., data = sr_train, method = "rpart", trControl = ctrl_dt,tuneLength = 25)
#prp(dtree$finalModel)
```

#### Random Forest
```{r}
ctrl_rf = trainControl(method = "oob")
rf = caret::train(risk_level ~., data = sr_train, method = "rf", trControl = ctrl_rf,
           tuneLength = 25, ntree = 250, importance = TRUE)
```

#### Neural Network
```{r}
train_nn<-sr_train
train_nn$risk_level = factor(train_nn$risk_level, levels = c("no","low","moderate","high"), labels = c(1,2,3,4))
dummy <- dummyVars(" ~ .", data=train_nn[c(1,3,4,6,7,9,10,13)])
newdata <- data.frame(predict(dummy, newdata = train_nn[c(1,3,4,6,7,9,10,13)]))
train_nn<-cbind(train_nn[c(2,5,8,11,12)],newdata)

nn <- keras_model_sequential() %>%  
  layer_flatten(input_shape = c(28)) %>%  
  layer_dense(units = 10, activation = "relu")%>%
  layer_dense(units = 6, activation = "relu")%>%
  layer_dense(units = 4, activation = "softmax")
nn%>% compile(
    loss = "categorical_crossentropy",
    optimizer = 'rmsprop',
    metrics = 'accuracy')
nn %>% fit(x = as.matrix(train_nn[,c(1:28)]), 
           y = as.matrix(train_nn[,c(29:32)]), 
          epochs = 15, validation_split = 0.3,verbose = 2)
```

### Testing

#### Predictions on Testing Set
```{r}
# KNN, DT, RF for prediction
pred_knn = predict(knn,sr_test)
pred_dtree = predict(dtree,sr_test)
pred_rf = predict(rf,sr_test)
```

```{r}
# Neural network for prediction
test_nn<-sr_test
test_nn$risk_level = factor(test_nn$risk_level, levels = c("no","low","moderate","high"), labels = c(1,2,3,4))
dummy1 <- dummyVars(" ~ .", data=test_nn[c(1,3,4,6,7,9,10,13)])
newdata1 <- data.frame(predict(dummy1, newdata = test_nn[c(1,3,4,6,7,9,10,13)]))
test_nn<-cbind(test_nn[c(2,5,8,11,12)],newdata1)

pred_nn = nn %>% predict(as.matrix(test_nn[,c(1:28)]))
pred_nn1 <- data.frame(pred_nn)
colnames(pred_nn1) <- c(1,2,3,4)
pred_nn1$risk_level <- apply(pred_nn1,1,which.max)
pred_nn1$risk_level = as.factor(pred_nn1$risk_level)
pred_nn1$risk_level <- factor(pred_nn1$risk_level, levels = c(levels(pred_nn1$risk_level), "1"))
```

```{r}
Evaluation<-data.frame(pred_knn, pred_dtree, pred_rf, pred_nn = pred_nn1$risk_level, Actual=sr_test$risk_level)
```


### True Positive Cases
```{r}
for (rows in Evaluation){
  correctM_dt = ifelse (pred_dtree == 3 & Evaluation$Actual == 3, 1, 0)
  correctH_dt = ifelse (pred_dtree == 4 & Evaluation$Actual == 4, 1, 0)
  correctL_dt = ifelse (pred_dtree == 2 & Evaluation$Actual == 2, 1, 0)
  
  correctM_rf = ifelse (pred_rf == 3 & Evaluation$Actual == 3, 1, 0)
  correctH_rf = ifelse (pred_rf == 4 & Evaluation$Actual == 4, 1, 0)
  correctL_rf = ifelse (pred_rf == 2 & Evaluation$Actual == 2, 1, 0)
  
  correctM_knn = ifelse (pred_knn == "moderate" & Evaluation$Actual == 3, 1, 0)
  correctH_knn = ifelse (pred_knn == "high" & Evaluation$Actual == 4, 1, 0)
  correctL_knn = ifelse (pred_knn == "low" & Evaluation$Actual == 2, 1, 0)
  
  correctM_nn = ifelse (Evaluation$pred_nn == 3 & Evaluation$Actual == 3, 1, 0)
  correctH_nn = ifelse (Evaluation$pred_nn == 4 & Evaluation$Actual == 4, 1, 0)
  correctL_nn = ifelse (Evaluation$pred_nn == 2 & Evaluation$Actual == 2, 1, 0)
}

sum(correctM_rf)
sum(correctH_rf)
sum(correctL_rf)

sum(correctM_nn)
sum(correctH_nn)
sum(correctL_nn)

sum(correctM_knn)
sum(correctH_knn)
sum(correctL_knn)

sum(correctM_dt)
sum(correctH_dt)
sum(correctL_dt)
table(Evaluation$Actual)

table_correct = data.frame(" " = c("**High**","**Moderate**","**Low**"), KNN = c(10,0,2574), 
                           Decision_Tree = c(82,0,2549), Random_Forest = c(68,0,2541),Neural_Network = c(25,1,2461),
                           Actual_val = c(412,335,2580))
kable(table_correct)
```

### F1-Score
```{r}
metrics_dtree <- ml_test(Evaluation$pred_dtree, Evaluation$Actual, output.as.table = TRUE)
metrics_rf <- ml_test(Evaluation$pred_rf, Evaluation$Actual, output.as.table = TRUE)
Evaluation$pred_knn = factor(Evaluation$pred_knn, levels = c("no","low","moderate","high"),labels = c(1,2,3,4))
metrics_knn <- ml_test(Evaluation$pred_knn, Evaluation$Actual, output.as.table = TRUE)

f1_dtree <- metrics_dtree$F1
f1_rf <- metrics_rf$F1
f1_knn <- metrics_knn$F1

#### NN
precision_h = 25/162
precision_m = 1/4
precision_l = 2461/3275

recall_h = 25/412
recall_m = 1/335
recall_l = 2461/2580

f1_nn_h = 2*(precision_h*recall_h)/(precision_h+recall_h)
f1_nn_m = 2*(precision_m*recall_m)/(precision_m+recall_m)
f1_nn_l = 2*(precision_l*recall_l)/(precision_l+recall_l)
f1_nn = c(f1_nn_l,f1_nn_m,f1_nn_h)

f1 = data.frame("." = c("Low","Moderate","High"), Decision_tree = f1_dtree[2:4],
                Random_forest = f1_rf[2:4],KNN = f1_knn[2:4], Neural_network = f1_nn)

kable(f1)
```

